Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-10 paper.tex
Add CUDA, because cross-API comparison is extremely deceiving. Revert grid stride Memcpy.

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans depart from early PRAM-like scans~\cite{Blelloch:1989:SAP} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers. Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory. Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{Merrill2009}. Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization. This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization. \thomas{I have written this paragraph like twenty times now, and it's still odious.} \raph{It's fine. Feel free to delete these notes in next round.}

@@

Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads. Recognizing that raking naturally aligns with this paradigm, Breitbart's method dynamically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels. Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup barrier without context switching. While global memory movement remains $O(3n)$, this consolidates the scan into a single kernel, minimizing launch overheads.

@@

\raph{I took out the statement about cache underutilization because I'm not sure it's true (L1 cache is shared by all subgroups in a multiprocessor) and I don't think it's essential to our argument. Also deleting John's comment; I don't think this is a case where we need to prove novelty over a previous publication, instead we just subsume it.}

\section{Decoupled Fallback}
\emph{Decoupled Fallback} addresses both inefficiencies. The key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially. To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is performed by a single thread.

\raph{This paragraph feels to me like it can be cut, but I won't insist on it.} Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to preclude deadlocks. \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking. While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible. Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, shifting rather than eliminating deadlock risks. Removing \emph{resource holding}---allowing workgroups to discard \emph{blocked} work tiles---is theoretically possible but strictly inferior to removing \emph{mutual exclusion}, as it incurs $t$ extra memory operations (discard-read-read instead of read).
\begin{algorithm}
  \small
  \SetAlgoLined
  \KwGlobalIn{%
    $\textit{in}[]$,
    $\textit{out}[]$,
    $\textit{spine}[]$,
    $\textit{tileBump}$,
  }
  \KwSharedIn{%
    $\textit{wg\_partials}[]$,
    $\textit{wg\_fallback}[]$,
    $\textit{wg\_prev\_reduction}$,
    $\textit{wg\_broadcast}$,
    $\textit{wg\_control}$,
  }
  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
  \KwConstants{Vectors per thread $\textit{p}$}

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models. \raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper". I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

@@

A key innovation of this paper is to \emph{split} the reduced values into multiple atomics, each carrying its own copy of the control flags. The message is received when relaxed atomic loads of all the parts agree on their control flag. This way, the payload is reconstructed intact, even in the absence of any guarantees from the memory model stronger than relaxed atomics.

@@

Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state. When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}. If the tile state is \emph{Inclusive} then the write must be skipped, especially because races in writes for the parts may leave the tile permanently in an unreadable state. Thus, a simple atomic store is not suitable. An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \texttt{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult). Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}. This allows an atomic \emph{maximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. Through testing, \john{across all devices (?)} we find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ \john{It would be worthwhile, I think, to run a test against CUB where the x axis is problem size and the y axis is throughput. A reviewer might think we're hiding something if we only report results for one (large) input size. I imagine that our performance will look similar to CUB's at all input sizes and that is the result we want.} and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \small
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
    \toprule
    Vendor & Device Name & FPG & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620       & \textbf{Yes} & 9   & 8--32  \\
    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\
    Apple  & M1 Pro Max             & \textbf{No}  & 64  & 32  \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32--64  \\
    \bottomrule
  \end{tabular*}
  \caption{Hardware Specifications of Testing Devices.}
  \label{tab:hardware}
\end{table}

\subsection{Performance Comparison}
To evaluate the performance of \emph{Decoupled Fallback} (\emph{CSDLDF}), we compare it against two key benchmarks: \emph{Memcopy}\textdagger{~\ref{sec:memcopy}} and \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}). \emph{RTS} serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique. \emph{Memcopy} is included as a point of reference, as it shares the same $2n$ global memory movement as \emph{Decoupled Fallback}, without any compute overhead and thus is guaranteed to be memory bandwidth bound. \john{Nit: I would use \texttt{memcpy} as the name of this benchmark to match how it's actually called in C etc.}

@@

\appendix
\section{Artifact}
\subsection{Baseline Comparisons: \emph{Memcopy}}
\label{sec:memcopy}
\begin{table}[h!]
  \small
  \centering
  \begin{tabular}{l r r}
    \toprule
    Device & \makecell{Memory \\ Bandwidth \\ (GB/s)} & \makecell{Expected \\ Throughput \\ (32-bit ele/sec)} \\
    \midrule
    Intel HD620 (Single-/Dual-Channel) & 17.1 / 34.1  & 2.13e9 / 4.26e9 \\
    ARM Mali-G78 MP20 / Pixel 6a (Tensor) & 51.2  & 6.40e9 \\
    Apple M1 Pro Max   & 200  & 25.0e9 \\
    Nvidia 2080 Super  & 448  & 56.0e9 \\
    AMD 7900 XT        & 800  & 100.0e9 \\
    \bottomrule
  \end{tabular}
  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second)}
  \label{tab:memory_bandwidth}
\end{table}
Although all backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcopy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation. It is well known that \emph{observed} peak memory bandwidth, tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above. We make a number of observations:
\begin{itemize}
  \item \textbf{Intel HD620:} As our HD620 laptop is only equipped with a single memory DIMM, its peak bandwidth is limited to the lower bound, 2.13e9 ele/sec. Thus, our observed bandwidth of 1.498e9 is reasonable.
  \item \textbf{ARM Mali-G78 MP20 / Pixel 6a (Tensor):} The observed throughput is significantly lower than expected, despite a vendor-listed bandwidth of 6.40e9 ele/sec. The cause of this discrepancy remains unclear, but we note other developers have observed similar speeds.
  \item \textbf{Apple M1 Pro Max:} Observed throughput matches expectations.
  \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
  \item \textbf{AMD 7900 XT:} Our above result, where \emph{CSDLDF} was observed to be faster than \emph{Memcopy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging. Furthermore, we see that our \emph{CSDLDF} throughput is within spec and reasonable.
\end{itemize}

>>>>

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans depart from early PRAM-like scans~\cite{Blelloch:1989:SAP} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers. Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory. Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{Merrill2009}. Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization. This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization.

@@

Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads. Recognizing that raking naturally aligns with this paradigm, Breitbart's method statically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels. Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup barrier without context switching. While global memory movement remains $O(3n)$, this consolidates the scan into a single kernel, minimizing launch overheads.

@@

\section{Decoupled Fallback}
\emph{Decoupled Fallback} addresses both inefficiencies. The key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially. To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is performed by a single thread\thomas{Is this clearer as "single subgroup?"}.

\begin{algorithm}
  \small
  \SetAlgoLined
  \KwGlobalIn{%
    $\textit{in}[]$,
    $\textit{out}[]$,
    $\textit{spine}[]$,
    $\textit{tileBump}$,
  }
  \KwSharedIn{%
    $\textit{wg\_partials}[]$,
    $\textit{wg\_fallback}[]$,
    $\textit{wg\_prev\_reduction}$,
    $\textit{wg\_broadcast}$,
    $\textit{wg\_control}$,
  }
  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
  \KwConstants{Vectors per thread $\textit{p}$}

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models~\cite{}. \raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper". I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

@@

A key innovation of this paper is to \emph{split} the reduced values into multiple atomics, each carrying its own copy of the tile state. The message is received when relaxed atomic loads of all the parts agree on their tile state. This way, the payload is reconstructed intact, even in the absence of any guarantees from the memory model stronger than relaxed atomics.

@@

Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state. When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}. If the tile state is \emph{Inclusive} then the write must be skipped, especially because races in writes for the parts may leave the tile permanently in an unreadable state. Thus, a simple atomic store is not suitable. An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \texttt{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult). Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}. This allows an \texttt{atomicMaximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. Through testing, \john{across all devices (?)} we find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ \john{It would be worthwhile, I think, to run a test against CUB where the x axis is problem size and the y axis is throughput. A reviewer might think we're hiding something if we only report results for one (large) input size. I imagine that our performance will look similar to CUB's at all input sizes and that is the result we want.} and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \small
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
    \toprule
    Vendor & Device Name & OBE? & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620       & \textbf{Yes} & 9   & 8--32  \\
    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\
    Apple  & M1 Pro Max             & \textbf{No}  & 64  & 32  \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32--64  \\
    \bottomrule
  \end{tabular*}
  \caption{Hardware Specifications of Testing Devices.}
  \label{tab:hardware}
\end{table}

\subsection{Performance Comparison}
\subsubsection{Performance Versus Contemporary Implementations:}
We first compare \emph{Decoupled Fallback} (\emph{CSDLDF}) to \emph{Decoupled Lookback} (\emph{CSDL}) in NVIDIA’s \emph{CUB} library\footnote{CUDA toolkit 12.8, CUB/CCCL VERSION}~\cite{}. As this test is limited to CUDA, evaluation is restricted to the 2080 Super. Due to WebGPU’s 128MB storage buffer limit, we test input sizes from $2^{10}$ to $2^{25}$\!.

To isolate the effect of \emph{Decoupled Fallback}, we implement our own \emph{CSDL} with an identical local-scan implementation. Our \emph{CSDL} implementation outperforms \emph{CUB} before reaching saturation, and likely due to this advantage, \emph{CSDLDF} achieves near-identical performance to \emph{CUB} despite its additional \emph{Chained-Propagation} overhead. While Dawn struggles, it still reaches saturation at larger inputs. However, cross-API comparisons remain challenging, as despite nearly identical high-level code, all implementations perform significantly better in CUDA, particularly for inputs smaller than $2^{20}$\!, where CUDA more effectively utilizes the 2080 Super’s 4MB L2 cache.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
\end{figure}

\subsubsection{Same-API Performance Versus Baseline Measures}
To evaluate the performance of \emph{Decoupled Fallback} (\emph{CSDLDF}), we compare it against two key benchmarks: \emph{Memcopy}\textdagger{~\ref{sec:memcopy}} and \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}). \emph{RTS} serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique. \emph{Memcopy} is included as a point of reference, as it shares the same $2n$ global memory movement as \emph{Decoupled Fallback}, without any compute overhead and thus is guaranteed to be memory bandwidth bound. \john{Nit: I would use \texttt{memcpy} as the name of this benchmark to match how it's actually called in C etc.}

@@

\clearpage
\appendix
\section{Artifact}
\subsection{Baseline Comparisons: \emph{Memcopy}}
\label{sec:memcopy}
\begin{table}[h!]
  \small
  \centering
  \begin{tabular}{l r r}
    \toprule
    Device & \makecell{Memory \\ Bandwidth \\ (GB/s)} & \makecell{Expected \\ Throughput \\ (32-bit ele/sec)} \\
    \midrule
    Intel HD620 (Single-/Dual-Channel) & 17.1 / 34.1  & 2.13e9 / 4.26e9 \\
    ARM Mali-G78 MP20 / Pixel 6a (Tensor) & 51.2  & 6.40e9 \\
    Apple M1 Pro Max   & 200  & 25.0e9 \\
    Nvidia 2080 Super  & 496  & 62.0e9 \\
    AMD 7900 XT        & 800  & 100.0e9 \\
    \bottomrule
  \end{tabular}
  \caption{Vendor-Specified Memory Bandwidth and Expected Throughput (32-bit Elements per Second)}
  \label{tab:memory_bandwidth}
\end{table}
Although all backend APIs support \emph{timestamp queries inside encoders}, Apple GPUs do not, so we are forced to implement our own \emph{Memcopy} kernel to measure memory bandwidth, as opposed to simply timing an API's copy buffer operation. It is well known that \emph{observed} peak memory bandwidth, tends to diverge from vendor-listed peak bandwidth, so for completeness we list these figures above. We make a number of observations:
\begin{itemize}
  \item \textbf{Intel HD620:} As our HD620 laptop is only equipped with a single memory DIMM, its peak bandwidth is limited to the lower bound, 2.13e9 ele/sec. Thus, our observed bandwidth of 1.498e9 is reasonable.
  \item \textbf{ARM Mali-G78 MP20 / Pixel 6a (Tensor):} The observed throughput is significantly lower than expected, despite a vendor-listed bandwidth of 6.40e9 ele/sec. The cause of this discrepancy remains unclear, but we note other developers have observed similar speeds.
  \item \textbf{Apple M1 Pro Max:} Observed throughput matches expectations.
  \item \textbf{Nvidia 2080 Super:} Observed throughput matches expectations.
  \item \textbf{AMD 7900 XT:} Our above result, where \emph{CSDLDF} was observed to be faster than \emph{Memcopy}, is likely due to compute latency counterintuitively improving memory access patterns, leading to reduced paging. Furthermore, we see that our \emph{CSDLDF} throughput is within spec and reasonable.
\end{itemize}
